---
title: "EDA-challenge"
author: "Nico Jaworski"
format: html
editor: visual
---

## Loading In Data

Here I am loading in the data-wrangling.csv file from a URL using the library {curl}

```{r}
#| message: false
#| warning: false
library(curl)
f <- curl("https://raw.githubusercontent.com/difiore/ada-datasets/main/data-wrangling.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```

## Establishing Variables

In this block of code I am creating 3 unique variables based on data in the data-wrangling csv. I am using the %\>% operator and the mutate function so that these variables are added to the dataset in R. The variables are BSD (body size dimorphism, ratio of average male to average female body mass), sex_ratio (number of adult females to adult males), and DI (defensibility index, ratio of day range length to home range diameter). I will then use names(d) to show that the variables have been added to the dataset

```{r}
#| message: false
#| warning: false
library(dplyr)
library(tidyverse)
d <- d %>%
  mutate(d, BSD = Body_mass_male_mean/Body_mass_female_mean) %>%
  mutate(d, sex_ratio = AdultFemale/AdultMales) %>%
  mutate(d, DI = DayLength_km/(2 * sqrt(HomeRange_km2/pi)))
names(d)
```

## Plotting the Data

In the following chunks of code, I am using the package {ggplot2} to create 3 different categories of plots (1) time spent moving and day range length, (2) group size and day range length, and (3) body size dimorphism and canine size dimorphism. Each category has 2 plots, one for all the primate species in the data and one which is split by family using the facet_wrap() function.

```{r}
#| message: false
#| warning: false
#plot the data
library(ggplot2)
library(dplyr)
#Movement and day length overall
ggplot(data = d, mapping = aes(x = Move, y = DayLength_km)) +
  geom_point(na.rm = TRUE) +
  ggtitle("Movement and Day Range Length Overall")

#Movement and day length by family
ggplot(data = d, mapping = aes(x = Move, y = DayLength_km)) +
  geom_point(na.rm = TRUE) +
  ggtitle("Movement and Day Range Length by Family") +
  facet_wrap(~Family)

#Group size and day range length overall
ggplot(data = d, mapping = aes(x = MeanGroupSize, y = DayLength_km)) +
  geom_point(na.rm = TRUE) +
  ggtitle("Group Size and Day Range Length Overall")

#Group size and day range by family
ggplot(data = d, mapping = aes(x = MeanGroupSize, y = DayLength_km)) +
  geom_point(na.rm = TRUE) +
  ggtitle("Group Size and Day Range Length by Family") +
  facet_wrap(~Family)


#Canine Dimorphism and Body Dimorphism Overall
ggplot(data = d, mapping = aes(x = BSD, y = Canine_Dimorphism)) +
  geom_point(na.rm = TRUE) +
  ggtitle("Body Size and Canine Size Dimorphism Overall")

#Canine Dimorphism and Body Dimorphism by Familt
ggplot(data = d, mapping = aes(x = BSD, y = Canine_Dimorphism)) +
  geom_point(na.rm = TRUE) +
  ggtitle("Body Size and Canine Size Dimorphism by Family") +
  facet_wrap(~Family)
```

## Creating a New Variable

Here I am creating a new variable called diet_strategy which will describe each species as a frugivore, folivore, or omnivore. I will then create a box plot showing the group size for each diet strategy.

```{r}
#| message: false
#| warning: false
library(tidyverse)
d <- d %>%
  mutate(d, diet_strategy = ifelse(Fruit >= 50, "frugivore", ifelse(Leaves >= 50,
  "folivore", ifelse(Fruit < 50 & Leaves < 50, "omnivore", NA))))

#now we are going to use box plots to plot 
ggplot(data = d, mapping = aes(x = diet_strategy, y = MeanGroupSize)) +
  geom_boxplot(na.rm = TRUE) +
  ggtitle("Diet Strategy and Mean Group Size")
```

## Arranging Family Average Brain Size

In this final chunk of code I will use the tools mutate(), select(), group_by(), summarize(), and arrange() from the package {dplyr}. This code will create a new variable called Binomial, trim the data frame down to just the variables Binomial, Family, Brain_size_species_mean, and Body_mass_male_mean. It will then group these variables by Family and determine the average value of Brain_size_species_mean and Body_mass_male_mean for each Family. These values will then be arranged in the data frame by increasing average brain size.

```{r}
#| message: false
#| warning: false
library(dplyr)
d <- d %>%
  mutate(d, Binomial = paste(Genus, Species, sep = " ")) %>%
  select(Binomial, Family, Brain_Size_Species_Mean, Body_mass_male_mean) %>%
  group_by(Family) %>%
  summarize(avgBrainSize = mean(Brain_Size_Species_Mean, na.rm = TRUE), avgBodyMassMale = mean(Body_mass_male_mean, na.rm = TRUE)) %>%
  arrange(avgBrainSize)
print(d)
```
